Procedural Terrain \\ {\large Final project for class CS6491 Computer Graphics}

### 2015-11-29 Paper/Paper.tex
improvements at the paper

\subsection{Generating the height map}
\subsubsection{Heightmap}
The next steps in the processing pipeline all require a heightmap. A height map is just a rectangular grid of values where each cell stores the height at this point. We also use the same datastructure to store moisture and temperature values.

Notation: Let $H$ be the heightmap. Then we identified the cell at position $i,j$ with either $H_{i,j}$ or $H[i,j]$, depending on the readability. When coordinates lie outside the map, we clamp them. When the coordinates are not integers, but lie between cells, we perform a bilinear interpolation between the neighboring cells. When two heightmaps, or in the general case two scalar- or vectorfields, are added or multiplied together in this paper, these are always element-wise operations.

@@

\section{Terrain Features}\label{TerrainFeatures}
Starting from the terrain from the previous chapter, the user has now the ability to fine-tune the terrain features.
This is an implementation of \cite{Hnaidi.2010} with some extensions. More examples and further explanations are available in this paper.

\begin{figure*}
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{images/diffusion1}
		\caption{feature curves}
		\label{fig:featurecurves}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{images/diffusion2}
		\caption{resulting terrain}
		\label{fig:resultingterrain}
	\end{subfigure}
	\caption{Defining feature curves}\label{fig:diffusion1}
\end{figure*}
\subsection{Feature curves}
In the center of this processing steps stand feature curves. Feature curves are added by the user and a solver (\ref{DiffusionSolver}) then modifies the terrain with respect to them.
Example feature curves forming hills can be seen in Fig. \ref{fig:featurecurves} (white spheres are the control points and the blue tupe is the interpolated curve) and the resulting terrain in Fig. \ref{fig:resultingterrain}.

@@

The gradient term is defined in the following way:
\begin{equation}
	G^{k+1}[i,j] = N^k[i,j]+G[i,j] \newline
\label{eq:Diffusion2}
\end{equation}
\begin{equation}
	N^k[i,j] = \textbf{n}_x^2 H^k[i-\text{sign}(\textbf{n}_x),j] + \textbf{n}_y^2 H^k[i, j-\text{sign}(\textbf{n}_y)]
\label{eq:Diffusion3}
\end{equation}
The gradient equation tend to satisfy the provided gradient $G$. $G$ is directly calculated from $\sin(\varphi_l)$ and $\sin(\varphi_r)$, i.e. from the slope angles defined at the control points.
The vector $\textbf{n}=(\textbf{n}_x,\textbf{n}_y)$ describes the normalized direction of the gradient. This vector is orthogonal to the feature curve when projected in the plane and point away from the curve. In Fig. \ref{fig:featsExist2}, some of them are displayed as gray lines going away from the plateau.

@@

\begin{figure*}
	\centering
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/diffusionEditExisting1}
		\caption{original terrain}
		\label{fig:featsExist1}
	\end{subfigure}
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/diffusionEditExisting2}
		\caption{adding feature curves}
		\label{fig:featsExist2}
	\end{subfigure}
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/diffusionEditExisting3}
		\caption{resulting terrain}
		\label{fig:featsExist3}
	\end{subfigure}
	\caption{Adding features to existing terrain}\label{fig:diffusion2}
\end{figure*}
\subsection{Integrate into existing terrain}
Defining the whole terrain only with the feature curves is very tedious. Therefore we want to start with an existing terrain, as described in chapter \ref{PolygonalMap}.
However, simply starting with the existing terrain as $H^0$ does not work because the Laplace term would just smooth out every feature apart from the feature curves.

@@

This case happens quite often as seen in Fig. \ref{fig:diffusion3}. Here we define a new feature curve right next to an existing mountain. Although the curve's slope does not touch the existing terrain, the smoothing term inside the diffusion solver also increases the height outside of the feature curve. This leads to a growth of the mountain on the left (the tip of the mountain is now much higher than the control point, before it was about the same height).
From a designer's perspective, this behavior might be not wanted because it is hard to predict and control.

@@

\subsection{Extension of the solver}
We propose now another solution that both keeps existing terrain features while it preserves the local control and removes the inferences mentioned before.
The idea is to limit the smoothing term $L$ to influence only a customizable region around a feature curve.
For this, we add two new parameters, $l_l$ and $l_r$, to each control point. They specify the local radius of an envelope around the plateau and slope of the feature curve.
It not only extends along the slope, but also over the start and end of the feature curve.

@@

Because we no limit the smoothing area, we can apply the solver directly on the original height map without smoothing out existing features. There is no need anymore to solve the diffusion equation on a separate map with relative heights as done in the previous section.

@@

Erosion is comes in many shapes in the nature: there is thermal erosion which decomposes larger stones into smaller ones, wind erosion which carries sand over long distances and finally water erosion.
Water erosion or hydraulic erosion is caused by flowing water that erodes the terrain, transports sediment along the river and finally deposits it somewhere.
Since the hydraulic erosion is the strongest one, we focus on this type of erosion in the presented framework.

@@

\subsection{Erosion Model}
The algorithms follows the following steps, which are repeatedly executed using the timestep $\Delta t$ (e.g. $\Delta t = 0.01$):
\begin{enumerate}
	\item increment water by rainfall
	\item flow simulation
	\item change water volume
	\item erosion, deposition and sediment transportation
	\item evaporation
\end{enumerate}
In the next sections, we use the following notation for describing the involved variables, all of them are scalarfields or vectorfields on the grid of the terrain:
\begin{itemize}
	\item terrain height $b$
	\item water height $d$
	\item suspended sediment $s$
	\item outflow flux $\textbf{f}=(f^L,f^R,f^T,f^B)$
	\item water velocity $\textbf{v}=(u,v)$
\end{itemize}
\begin{center}
	\includegraphics[width=0.4\textwidth]{images/erosionVariables}
\end{center}

@@

\subsubsection{Water volume change}\label{Erosion3}
After computing the outflow flux, we can define the change of the water volume as:
\begin{equation}
\begin{array}{rcl}
	\Delta V[x,y] &=& \Delta t \cdot (\sum{f_{in}} - \sum{f_{out}}) \\
	              &=& \Delta t \cdot (f_{t+\Delta t}^L[x+1,y] + f_{t+\Delta t}^R[x-1,y] \\
								& & f_{t+\Delta t}^B[x,y+1] + f_{t+\Delta t}^T[x,y-1] \\
								& & - \sum_{i\in\{L,R,T,B\}}{f_{t+\Delta t}^i[x,y]} )
\end{array}
\label{eq:ErosionVolume1}
\end{equation}
The water height per cell is then modified using:
\begin{equation}
	d_{t+\Delta t}[x,y] = d_t[x,y] + \Delta V[x,y]
\label{eq:ErosionVolume2}
\end{equation}
For the next step, \ref{ErosionDeposition}, we need to compute the horizontal velocity of the water $\textbf{v}=(u,v)$.
Let $\bar{d}$ be the arithmetic average of the water height before and after equation \ref{eq:ErosionVolume2}.
Then the velocity along the x-axis is compute as:
\begin{equation}
	u = \frac{f^R[x-1,y]-f^L[x,y] + f^R[x,y]-f^L[x+1,y]}{2\bar{d}}
\label{eq:ErosionVolume3}
\end{equation}
$v$, the velocity along the y-axis is computed similarly.

@@

Fig. \ref{fig:erosion} shows the simulation on a larger scale. The original terrain is shown in \ref{fig:erosion1}, then in \ref{fig:erosion2}, the simulation is running. The displayed terrain height is the terrain height plus the water height, so lakes, waves and so on can be seen. The water height also specifies how much "blue" is drawn on top of the terrain. The terrain is textured with grass until a specific slope angle is reached, then stone is used (with blending between them). By that, you can get a feeling of the local slope angle. In \ref{fig:erosion3}, you can see the difference from the original terrain to the new terrain. White indicates depositions, black stands for erosion. Finally, \ref{fig:erosion4} shows the final terrain without water. Especially note the change in the coast line and that the terrain is much more "bumpy", more stone is used in the texturing of the terrain.

>>>>

\subsection{Generating the height map}
\subsubsection{Heightmap}
The next steps in the processing pipeline all require a height map. A height map is just a rectangular grid of values where each cell stores the height at this point. We also use the same datastructure to store moisture and temperature values.

Notation: Let $H$ be the heightmap. Then we identified the cell at position $i,j$ with either $H_{i,j}$ or $H[i,j]$, depending on the readability. When coordinates lie outside the map, we clamp them. When the coordinates are not integers, but lie between cells, we perform a bilinear interpolation between the neighboring cells. When two heightmaps, or in the general case two scalar- or vectorfields, are added or multiplied \mbox{together} in this paper, these are always element-wise operations.

@@

\section{Terrain Features}\label{TerrainFeatures}
Starting from the terrain from the previous chapter, the user now has the ability to fine-tune the terrain features.
This is an implementation of \cite{Hnaidi.2010} with some extensions. More examples and further explanations are available in this paper.

\begin{figure*}
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{images/diffusion1}
		\caption{feature curves}
		\label{fig:featurecurves}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{images/diffusion2}
		\caption{resulting terrain}
		\label{fig:resultingterrain}
	\end{subfigure}
	\caption{Defining feature curves}\label{fig:diffusion1}
\end{figure*}
\subsection{Feature curves}
In the center of this processing step stand feature curves. Feature curves are added by the user and a solver (\ref{DiffusionSolver}) then modifies the terrain with respect to them.
Example feature curves forming hills can be seen in Fig. \ref{fig:featurecurves} (white spheres are the control points and the blue tupe is the interpolated curve) and the resulting terrain in Fig. \ref{fig:resultingterrain}.

@@

The gradient term is defined in the following way:
\begin{equation}
	G^{k+1}[i,j] = N^k[i,j]+G[i,j] \newline
\label{eq:Diffusion2}
\end{equation}
\begin{equation}
	N^k[i,j] = \textbf{n}_x^2 H^k[i-\text{sign}(\textbf{n}_x),j] + \textbf{n}_y^2 H^k[i, j-\text{sign}(\textbf{n}_y)]
\label{eq:Diffusion3}
\end{equation}
The gradient equation tends to satisfy the provided gradient $G$. $G$ is directly calculated from $\sin(\varphi_l)$ and $\sin(\varphi_r)$, i.e. from the slope angles defined at the control points.
The vector $\textbf{n}=(\textbf{n}_x,\textbf{n}_y)$ describes the normalized direction of the gradient. This vector is orthogonal to the feature curve when projected in the plane and point away from the curve. In Fig. \ref{fig:featsExist2}, some of them are displayed as gray lines going away from the plateau.

@@

\begin{figure*}
	\centering
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/diffusionEditExisting1}
		\caption{original terrain}
		\label{fig:featsExist1}
	\end{subfigure}
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/diffusionEditExisting2}
		\caption{adding feature curves}
		\label{fig:featsExist2}
	\end{subfigure}
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/diffusionEditExisting3}
		\caption{resulting terrain}
		\label{fig:featsExist3}
	\end{subfigure}
	\caption{Adding features to existing terrain}\label{fig:diffusion2}
\end{figure*}
\subsection{Integration into existing terrain}
Defining the whole terrain only with the feature curves is very tedious. Therefore we want to start with an existing terrain, as described in chapter \ref{PolygonalMap}.
However, simply starting with the existing terrain as $H^0$ does not work because the Laplace term would just smooth out every feature apart from the feature curves.

@@

This case happens quite often as seen in Fig. \ref{fig:diffusion3}. Here we define a new feature curve right next to an existing mountain. Although the curve's slope does not touch the existing terrain, the smoothing term inside the diffusion solver also increases the height outside of the feature curve. This leads to a growth of the mountain on the left (the tip of the mountain is now much higher than the control point, before it was about the same height).
From a designer's perspective, this behavior might not be desired because it is hard to predict and control.

@@

\subsection{Extension of the solver}
We propose now another solution that both keeps existing terrain features while it preserves the local control and removes the inferences mentioned before.
The idea is to limit the smoothing term $L$ to influence only a customizable region around a feature curve.
For this, we add two new parameters, $l_l$ and $l_r$, to each control point. They specify the local radius of an envelope around the plateau and slope of the feature curve.

@@

Because we no limit the smoothing area, we can apply the solver directly on the original height map without smoothing out existing features. There is no need any more to solve the diffusion equation on a separate map with relative heights as done in the previous section.

@@

Erosion comes in many shapes in the nature: there is thermal erosion which decomposes larger stones into smaller ones, wind erosion which carries sand over long distances and finally water erosion.
Water erosion or hydraulic erosion is caused by flowing water that erodes the terrain, transports sediment along the river and finally deposits it somewhere.
Since the hydraulic erosion is the strongest one, we focus on this type of erosion in the presented framework.

@@

\subsection{Erosion Model}
The algorithms follows the following steps, which are repeatedly executed using the timestep $\Delta t$ (e.g. $\Delta t = 0.01$):
\begin{enumerate}
	\item water increment by rainfall
	\item flow simulation
	\item water volume change
	\item erosion, deposition and sediment transportation
	\item evaporation
\end{enumerate}
In the next sections, we use the following notation for describing the involved variables, all of them are scalarfields or vectorfields on the grid of the terrain:
\begin{itemize}
	\item terrain height $b$
	\item water height $d$
	\item suspended sediment $s$
	\item outflow flux $\textbf{f}=(f^L,f^R,f^T,f^B)$
	\item water velocity $\textbf{v}=(u,v)$
\end{itemize}
\begin{center}
	\includegraphics[width=0.4\textwidth]{images/erosionVariables}
\end{center}

@@

\subsubsection{Water volume change}\label{Erosion3}
After computing the outflow flux, we can define the change of the water volume as:
\begin{equation}
\begin{array}{rcl}
	\Delta V[x,y] &=& \Delta t \cdot (\sum{f_{in}} - \sum{f_{out}}) \\
	              &=& \Delta t \cdot (f_{t+\Delta t}^L[x+1,y] + f_{t+\Delta t}^R[x-1,y] \\
								& & f_{t+\Delta t}^B[x,y+1] + f_{t+\Delta t}^T[x,y-1] \\
								& & - \sum_{i\in\{L,R,T,B\}}{f_{t+\Delta t}^i[x,y]} )
\end{array}
\label{eq:ErosionVolume1}
\end{equation}
The water height per cell is then modified using:
\begin{equation}
	d_{t+\Delta t}[x,y] = d_t[x,y] + \Delta V[x,y]
\label{eq:ErosionVolume2}
\end{equation}
For the next step, \ref{ErosionDeposition}, we need to compute the horizontal velocity of the water $\textbf{v}=(u,v)$.
Let $\bar{d}$ be the arithmetic average of the water height before and after equation \ref{eq:ErosionVolume2}.
Then the velocity along the x-axis is computed as:
\begin{equation}
	u = \frac{f^R[x-1,y]-f^L[x,y] + f^R[x,y]-f^L[x+1,y]}{2\bar{d}}
\label{eq:ErosionVolume3}
\end{equation}
$v$, the velocity along the y-axis is computed similarly.

@@

Fig. \ref{fig:erosion} shows the simulation on a larger scale. The original terrain is shown in \ref{fig:erosion1}, then in \ref{fig:erosion2}, the simulation is running. The displayed terrain height is the terrain height plus the water height, so lakes, waves and so on can be seen. The water height also specifies how much "blue" is drawn on top of the terrain. The terrain is textured with grass until a specific slope angle is reached, then stone is used (with blending between them). By that, you can get a feeling of the local slope angle. In \ref{fig:erosion3}, you can see the difference from the original terrain to the new terrain. White indicates depositions, black stands for erosion. Finally, \ref{fig:erosion4} shows the final terrain without water. Especially note the change in the coast line and that the terrain is much more "bumpy", the stone texture is used more often.
